An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-11-17 paper/main.tex
weiter

A typical neuron in a neural network performs a weighted sum of its inputs,
followed by an activation function (depicted in~\eqref{eq:activation}):

\begin{equation}\label{eq:activation}
	a=\omega^\top x+b,\quad\hat{y}=\sigma(a)
\end{equation}

where $\omega \in \mathbb{R}^n$ is the weight vector, $b \in \mathbb{R}$ is the
bias, $x \in \mathbb{R}^n$ is the input, and $\sigma$ is a nonlinear activation
function (e.g., sigmoid or ReLU). Training is conducted via backpropagation,
where gradients of a loss function $L(\hat{y}, y)$ with respect to the model
parameters are computed. For example, using the mean squared error:

\begin{align}\label{eq:mse}
	L(f(x_1, x_2)) & = {(\sigma(\omega^\top x + b) - y)}^2 \nonumber \\
	               & = {(\sigma(a) - y)}^2 \nonumber                 \\
	               & = {(\hat y - y)}^2
\end{align}
\begin{alignat}{3}\label{eq:grad-w1}
	\pdv{L}{\omega_1} & = \pdv{L}{\sigma} &  & \pdv{\sigma}{a}      &  & \pdv{a}{\omega_1} \nonumber \\
	                  & = 2(\hat y - y)~~ &  & \odv{\sigma(a)}{a}~~ &  & x_1
\end{alignat}

Memory and computational requirements during training are substantial, as
gradients, activations, and parameters must be stored and updated iteratively.
In contrast, inference requires only a forward pass, making it more amenable to
optimization via precision reduction.

>>>>

A typical neuron in a neural network performs a weighted sum of its inputs,
followed by an activation function which results in the neuron output $h$
depicted in~\eqref{eq:activation}. Here, $\omega\in\mathbb{R}^n$ is the weight
vector, $b \in \mathbb{R}$ is the bias, $x \in \mathbb{R}^n$ is the input, and
$\sigma$ is a nonlinear activation function (e.g., sigmoid or ReLU).

\begin{equation}\label{eq:activation}
	a=\omega^\top x+b,\quad h=\sigma(a)
\end{equation}

\noindent
In our MLP, hidden layers use the Rectified Linear Unit (ReLU), so
$\sigma=\mathrm{ReLU}$, while the output layer is linear, producing raw scores
(also called logits) $z$ later used in~\eqref{eq:ce}. ReLU~\eqref{eq:relu}
keeps positive activations linear and sets negatives to zero. An additional
benefit is that the computation is very fast and simple compared to more
complex activation functions.

\begin{equation}\label{eq:relu}
	\mathrm{ReLU}(a)=\max(0,a),\qquad\frac{d}{da}\mathrm{ReLU}(a)=\mathbbm{1}\{a>0\}
\end{equation}

\noindent
Training of the network is conducted via backpropagation, where gradients of a
loss function $L$ are computed with respect to the model parameters. For
multi-class classification with $C=10$ classes, the last layer outputs logits
$z\in\mathbb{R}^C$. To convert the logits into probabilities, usually softmax
defined as $p_k=\frac{e^{z_k}}{\sum_{j=1}^C e^{z_j}}$ is used. The probability
of the true class $y$ is then $p_y$. In our experiments, though, we train
directly on the logits using a simplified version of the cross-entropy loss
(the negative log-likelihood of a softmax classifier) implemented in
Pytorch\footnote{\url{https://pytorch.org/}}:

\begin{equation}\label{eq:ce}
	L(z,y)=-\log p_y=-z_y+\log\sum_{j=1}^C e^{z_j},
\end{equation}

whose gradients are

\begin{equation}\label{eq:cegrads}
	\frac{\partial L}{\partial z_k}=p_k-\mathbbm{1}\{k=y\},\qquad\frac{\partial L}{\partial W}=(p-e_y)h^\top,\qquad\frac{\partial L}{\partial b}=p-e_y,
\end{equation}

with $W$ and $b$ the weight matrix and bias of the last linear layer, $h$ the
layer input, and $e_y$ the one-hot vector of the target class. Here,
$\mathbbm{1}\{k=y\}$ is the indicator: it equals $1$ if $k=y$ and $0$
otherwise.

Memory and computational requirements during training are substantial, as
weights, gradients, activations, and parameters must be stored and updated
iteratively. Training and inference can both be optimized using precision
reduction and related techniques, but inference is more well suited to
optimizations because it only requires a forward pass.
